---
title: "Premimum Bond Prize Simulator"
format: 
  dashboard:
    scrolling: true
    theme: sandstone
    nav-buttons:
      - icon: github
        href: https://github.com/drclongstaff/premium_bonds
      - icon: bluesky
        href: http://bsky.app/profile/drclongstaff.bsky.social
      - icon: globe2
        href: https://drclongstaff.github.io/shiny-clots
      - text: Contact
        href: mailto:drclongstaff@gmail.com
server: shiny
---

```{r}
#| context: setup
library(ggplot2)
library(plotly)
library(gt)
library(bslib)
library(data.table)
# Read once at startup, not on every input change
allMonths <- c("Jan", "Feb", "March", "April", "May", "June", "July", "Aug", "Sept", "Oct", "Nov", "Dec")
# set up the pool for prize draws = 1 in 23000 zeros
prizePool <- rep(0, 23000)
x <- round(runif(1) * 23000)
prizePool[x] <- 1

prizeVal <- c(1e6, 100000, 50000, 25000, 10000, 5000, 1000, 500, 100, 50, 25)
prizeNos <- c(2, 71, 142, 255, 712, 1421, 15021, 45063, 1535556, 1535556, 2803175)
allPrizes <- sum(prizeNos)
prizeProb <- prizeNos / allPrizes
```

# Are premium bonds a worthwhile investment? Simulate winnings for £ value of bonds held.

```{r}
#| context: server
# allMonths <- c("Jan", "Feb", "March", "April", "May", "June", "July", "Aug", "Sept", "Oct", "Nov", "Dec")
Earnest <- function(stake, pool, vals, probs) {
  mywins <- sample(pool, stake, replace = TRUE)
  monthPrizes <- sum(mywins)
  monthWinnings <- sample(vals, monthPrizes, replace = TRUE, prob = probs)
  monthTotal <- sum(monthWinnings)
  yearlyReturn <- round((12 * monthTotal / stake) * 100, 2)

  # Collapse winnings into a readable string
  winningsStr <- if (length(monthWinnings) == 0) "0" else paste(sort(monthWinnings), collapse = ", ")
  results <- c(monthPrizes, monthTotal, yearlyReturn, winningsStr)
}

bondStake <- reactive({
  input$go
  bondStake <- input$bondStake
})

yearRes <- reactive({
  bondStake <- bondStake()
  yearRes <- data.frame(t(sapply(allMonths, function(x) Earnest(bondStake, prizePool, prizeVal, prizeProb))))
  colnames(yearRes)[c(1, 2, 3, 4)] <- c("Prizes", "Value", "Yield", "Wins")

  # add a summary row to the table
  totalsRow <- data.frame(
    # Month = " ",
    Prizes = sum(as.numeric(yearRes$Prizes)),
    Value = sum(as.numeric(yearRes$Value)),
    Yield = round(100 * (sum(as.numeric(yearRes$Value)) / input$bondStake), 2),
    Wins = " "
    # Prizes = paste(sort(unlist(strsplit(yearRes$Prizes, ", "))), collapse = ", ")
  )

  yearRes <- rbind(yearRes, totalsRow)
  yearRes <- cbind("Month" = c(allMonths, "Total"), yearRes)
  yearRes
  
})
output$yearResTab <- render_gt({
  yearRes <- yearRes() |>
    gt() |>
    tab_header(
      title = md("Simulated prizes for one year of draws"),
      subtitle = paste0("for £", input$bondStake, " bonds held")
    ) |>
    opt_table_font(stack = "humanist") |>
    opt_stylize(style = 2, color = "green")
  yearRes
})

# Long term simulations
earnMonth <- function(stake, pool, vals, probs) {
  mywins <- sample(pool, stake, replace = TRUE)
  monthPrizes <- sum(mywins)
  monthWinnings <- sample(vals, monthPrizes, replace = TRUE, prob = probs)
  monthWin <- sum(monthWinnings)
}

earnYear <- function(stake, pool, vals, probs) {
  Resmnth <- sapply(1:12, function(x) earnMonth(stake, pool, vals, probs))
  Resyrtotal <- sum(Resmnth)
}

simYrs <- reactive({
  input$do
  simYrs <- input$simYrs
})

ResLife <- reactive({
  ResLife <- sapply(1:simYrs(), function(x) earnYear(input$bondStake, prizePool, prizeVal, prizeProb))
  ResLife <- sort(ResLife, decreasing = TRUE)
})

output$lifeSim <- renderPlotly({
  Win <- ResLife()
  meanResAll <- round(mean(Win), 2)
  meanResYield <- round((100 * meanResAll / input$bondStake), 2)
  forText <- simYrs() / 8
  p <- ggplot() +
    aes(Win) +
    geom_histogram(binwidth = 0.075, fill = "steelblue", col = "lightblue", aes(text = paste("win=", Win))) +
    scale_x_continuous(trans = "log10") +
    geom_vline(xintercept = meanResAll, col = "pink", lwd = 1, lty = "dashed") +
    annotate("text", x = meanResAll, y = -forText * 0.1, label = paste0(meanResYield, "%")) +
    labs(title = paste("Return from ", input$simYrs, "years of investment")) +
    xlab("£ Value of prizes") +
    theme_minimal()
  ggplotly(p, tooltip = "text")
})

```

## Row {height = 15% .panel-input}

```{r}
#| layout-ncol: 3
helpText(h5("Adjust the £ value of bonds held in the green box and years of simulation in the blue box. Click on the green or blue boxes to update the simulation."))

actionButton("go",
  numericInput("bondStake", "Bonds £",
    value = 50000, min = 1, max = 50000, step = 1000, width = "50%",
    updateOn = "blur"
  ),
  style = "color: #1B5E20; background-color: #DCEDC8; border-color: #F4511E"
)

actionButton("do",
  numericInput("simYrs", "Years of simulation", value = 100, min = 1, max = 1000, step = 10, width = "50%"),
  style = "color: #303F9F; background-color: #BBDEFB; border-color: #000000"
) 
```

## Row {.fill height="85%"}

### Column {.tabset}

#### Table

```{r}
# Summary table of data
probTab <- data.frame(
  "Prize" = LETTERS[1:11],
  "Value" = prizeVal,
  "Number" = prizeNos,
  "Probability" = signif(prizeProb, 3)
)

summaryRow <- data.frame("Prize" = "Totals", "Value" = sum(prizeVal), "Number" = allPrizes, "Probability" = round(sum(prizeProb)))

summaryTab <- rbind(probTab, summaryRow)

summaryTab |>
  gt() |>
  opt_stylize(style = 2, color = "pink") |>
  opt_table_font(stack = "humanist") |>
  tab_header(
    title = md("Premium bond prizes available")
  )
```

These are the number and value of available prizes each month. The probablility column is calculated from the number of each prize and total number. Other data is taken from the NS&I [website](https://www.nsandi.com/products/premium-bonds "NS&I background")

#### Background

From April 2026 the chances of winning each month are 1/23000 for each £1 bond. The prizes available and the associated probabilities of winning are shown in the left table.

Calculating the longterm winnings from premimum bonds is achieved in 2 steps.

1.  A vector containing 22999 'zeros' and a randomly placed single 'one' is generated. This is sampled m times, where m is the £ value of bonds held. The resulting vector of length m is summed to give the number of prizes won, n, for that month.

2.  The number of prizes generated from step 1, n, is used in a new sampling function where prizes shown in the lefthand table are sampled n times taking into account their associated probabilities.

The output over a year is presented in the table in the middle panel showing the number of prizes won and the values, along with a summary for the year.

The right panel is for simulating winnings over a number of chosen years. The histogram presents the distribution of prizes. This distribution is approximately normal on a log scale, with a few outlying prizes of high value. The log scale is not surprising (see more details [here](https://connect.posit.cloud/drclongstaff/content/019cf824-922f-b4ad-2c7f-fff03cb7ad5e "Premium bond prize finder")).

New simulations are achieved by adjusting the input numbers for the value of bonds held and the years of simulation, then clicking on the green and blue boxes.

### Column

#### Table

```{r}

gt_output("yearResTab")

```

The simulated prizes over a year are shown here, including individual prizes each month and a summary for the year. The yield column is the % return on investment scaled to a full year.

### Column

#### Plot

```{r}

plotlyOutput("lifeSim")

```

The plot shows the distribution of wins each year for the specified number of years. The pink line indicates the % return which should be around 3.3% according to the NS&I [website](https://www.nsandi.com/products/premium-bonds "NS&I premium bonds"). Payments are tax-free, which may be important for higher rate tax payers.
