---
title: "Premimum Bond Prize Simulator"
format: 
  dashboard:
    scrolling: false
    theme: materia
    nav-buttons:
      - icon: github
        href: https://github.com/drclongstaff/premium_bonds
      - icon: bluesky
        href: http://bsky.app/profile/drclongstaff.bsky.social
      - icon: globe2
        href: https://drclongstaff.github.io/shiny-clots
      - text: Contact
        href: mailto:drclongstaff@gmail.com
server: shiny
---

```{r}
#| context: setup
library(ggplot2)
library(plotly)
library(gt)
library(bslib)
library(data.table)
# Read once at startup, not on every input change
allMonths <- c("Jan", "Feb", "March", "April", "May", "June", "July", "Aug", "Sept", "Oct", "Nov", "Dec")
#set up the pool for prize draws = 1 in 23000 zeros
prizePool <- rep(0, 23000)
x <- round(runif(1)*23000)
prizePool[x] <- 1

prizeVal <- c(1e6, 100000, 50000, 25000, 10000, 5000, 1000, 500, 100, 50, 25)
prizeNos <- c(2,71,142,255,712,1421,15021, 45063, 1535556, 1535556, 2803175)
allPrizes <- sum(prizeNos)
prizeProb <- prizeNos/allPrizes
```



# Are premium bonds a worthwhile investment? Simulate winnings for £ bonds held.

```{r}
#| context: server
#allMonths <- c("Jan", "Feb", "March", "April", "May", "June", "July", "Aug", "Sept", "Oct", "Nov", "Dec")
Earnest <- function(stake, pool, vals, probs){
  
  mywins <- sample(pool,stake, replace = TRUE)
  monthPrizes <- sum(mywins)
  monthWinnings <- sample(vals, monthPrizes, replace = TRUE,  prob = probs)
  monthTotal <- sum(monthWinnings)
  yearlyReturn <- round((12*monthTotal/stake)*100, 2)
  
  # Collapse winnings into a readable string
  winningsStr <- if (length(monthWinnings) == 0) "0" else paste(sort(monthWinnings), collapse = ", ")
  results <- c( monthPrizes, monthTotal, yearlyReturn, winningsStr)
  
}

bondStake<- reactive({
  input$go
  bondStake <- input$bondStake
  
})

yearRes <- reactive({
  bondStake <- bondStake()
  yearRes <- data.frame(t(sapply(allMonths, function(x) Earnest(bondStake, prizePool, prizeVal, prizeProb))))
colnames(yearRes)[c(1,2,3,4)] <- c( "Prizes", "Value", "Yield", "Wins")

#add a summary row to the table
totalsRow <- data.frame(
  #Month = " ",
  Prizes      = sum(as.numeric(yearRes$Prizes)),
  Value  = sum(as.numeric(yearRes$Value)), 
  Yield = round(100*(sum(as.numeric(yearRes$Value))/input$bondStake), 2),
  Wins = " "
  #Prizes = paste(sort(unlist(strsplit(yearRes$Prizes, ", "))), collapse = ", ")
  
)

#yearRes <- cbind("Month"=allMonths, yearRes)
yearRes <- rbind(yearRes, totalsRow)
yearRes <- cbind("Month"=c(allMonths, "Total"), yearRes)
#rownames(yearRes)[nrow(yearRes)] <- "Total"

yearRes

})
output$yearResTab <- render_gt({
  yearRes <- yearRes() |> 
    gt() |> 
    opt_stylize(style = 2, color = "green")  
  yearRes
  
})

#Long term simulations
earnYear <- function(stake, pool, vals, probs){
  
  earnMonth <- function(stake, pool, vals, probs){
    mywins <- sample(pool,stake, replace = TRUE)
    monthPrizes <- sum(mywins)
    monthWinnings <- sample(vals, monthPrizes, replace = TRUE,  prob = probs)
    monthWin <- sum(monthWinnings)
  }
  
  Resmnth <- sapply(1:12, function(x) earnMonth(stake, pool, vals, probs))
  Resyrtotal <- sum(Resmnth)
}

simYrs<- reactive({
  input$do
  simYrs <- input$simYrs
  
})

ResLife <- reactive({
  
  ResLife <- sapply(1:simYrs(), function(x) earnYear(input$bondStake, prizePool, prizeVal, prizeProb))
  ResLife <- sort(ResLife, decreasing = TRUE)

})

output$lifeSim <- renderPlotly({
#hist(ResLife())
  Win <- ResLife()
  meanResAll <- round(mean(Win),2)
  meanResYield <- round((100*meanResAll/input$bondStake),2)
  forText <- simYrs()/8
  p <- ggplot() + aes(Win)+
  geom_histogram(binwidth = 0.075, fill="steelblue", col = "white", aes(text=paste("win=", Win)))+
  scale_x_continuous(trans='log10')+
  geom_vline(xintercept = meanResAll, col = "pink", lwd = 1, lty = "dashed")+
  annotate("text", x=meanResAll, y=-forText*0.1, label=paste0(meanResYield, "%"))+
  labs(title = paste("Return from ", input$simYrs, "years of investment"))+
  xlab("£ Value of prizes")+
  theme_minimal()
  ggplotly(p, tooltip = "text")
  
  #p |> layout(hoverinfo = 10^x)
  #p+add_trace(x = ResLife(),  marker = list(size = 10, color = "pink", line = list(color = "red", width = 2)), hoverinfo = x)
})


```


## Row {height = 20% .panel-input}

```{r}
#| layout-ncol: 3
#helpText(h5("Change bonds held or years of simulation"))
#numericInput("bondStake","Bonds £", value = 50000, min=1, max=50000, width = "30%")

helpText(h5("Adjust the £ of bond holding and years for long term simulation"))

actionButton("go", 
             numericInput("bondStake","Bonds £", value = 50000, min=1, max=50000, width ="50%"),
             icon("Refresh"), style="color: #1B5E20; background-color: #DCEDC8; border-color: #F4511E")

actionButton("do",
              numericInput("simYrs", "Years of simulation", value=100, min=1, max=1000, width = "50%"),
              icon("Refresh"), style="color: #303F9F; background-color: #BBDEFB; border-color: #000000")
    #F8BBD0
```


## Row {height="80%" .fill}

### Column {.tabset}

#### Table

```{r}
#Summary table of data

probTab <- data.frame("Prize"=LETTERS[1:11],
                      "Value"=prizeVal,
                      "Number"=prizeNos,
                      "Probability"=signif(prizeProb,3)
)

summaryRow <- data.frame("Prize"="Totals", "Value"= sum(prizeVal), "Number"=allPrizes, "Probability" = round(sum(prizeProb)))

summaryTab <- rbind(probTab, summaryRow)

summaryTab |> gt() |> 
  #cols_label(Prize ~ "Prize Value", Number ~ "Number of Prizes", Total_value ~ "Total Value") |> 
  #tab_header(title = "Expected prizes") |> 
  opt_stylize(style = 2, color = "pink") |> 
  opt_table_font(stack = "humanist") #|> 
  #tab_footnote(
    #footnote = "There are 2,608,195 unclaimed prizes totalling £108,147,675"
    
  #)
  
```

#### Background

From April 2026 the chances of winning each month are 1/23000 for each £1 bond. The prizes available and the associated probabilities of winning are shown in the left table.

Calculating the longterm winnings from premimum bonds is achieved in 2 steps.

1. A vector containing 22999 'zeros' and a randomly placed single 'one' is generated. This is sampled m times, where m is the £ value of bonds held. The resulting vector of length m is summed to give the number of prizes won, n, for that month.

2. The number of prizes generated from step 1, n, is used in a new sampling function. Here prizes shown in the table are sampled n times with the associated probabilites shown in the table.

The output over a year is presented in the table in the middle panel showing the number of prizes won and the values, along with a summary for the year.

The right panel is for simulating winnings over a number of chosen years. The histogram presents a distribution of prizes and the % yield based on overall winnings and bond holdings. 

New simulations are achieved by adjusting the input numbers for the value of bonds held and the years of simulation, then clicking on the green and blue boxes.


### Table


```{r}
scrolling = FALSE
#numericInput("bondStake",NULL, value = 50000, min=1, max=50000, width = "20%", updateOn = "blur")
#actionButton("go", 
           #  numericInput("bondStake","Refresh", value = 50000, min=1, max=50000, width = "50%"), 
           #  icon("Refresh"), class = "btn-primary", width = "50%")
gt_output("yearResTab")
```

### Column {.tabset}

#### Plot

```{r}
#numericInput("simYrs", NULL, value=100, min=1, max=1000, width = "100%", updateOn = "blur")

plotlyOutput("lifeSim")
```


#### Summary

